---
title: "Homework 2: The Big Short"
output: html_document
---
 
**This homework is due Friday February 26, 2016 at 5:00 PM. When complete, submit your code in the R Markdown file and the knitted HTML via GitHub.**

# Background 

This homework is motivated by circumstances surrounding the [financial crisis of 2007-2008](https://en.wikipedia.org/wiki/Financial_crisis_of_2007%E2%80%9308). We titled the homework _The Big Short_, after the book on the same topic that was also recently made into a movie.

Part of what caused the financial crisis was that the risk of certain [securities](https://en.wikipedia.org/wiki/Security_(finance)) sold by financial institutions were  underestimated. Specifically, 
the risk of mortgage-backed securities (MBS) and collateralized debt obligations (CDO), the price of which depends on homeowners making their monthly payments, was grossly underestimated. A combination of factors resulted in many more defaults than were expected. This resulted in a crash of the prices of these securities. As a consequence, banks lost so much money that they needed bailouts to avoid default.

Here we present a **very** simplified version of what happened with some of these securities. Hopefully it will help you understand how a wrong assumption about the statistical behavior of events can lead to substantial differences between what the model predicts and what actually happens. Specifically, we will see how using an independence assumption can result in misleading conclusions. Before we start with the specific application we ask you about a simple casino game.

# Problem 1

In the game of [roullete](https://en.wikipedia.org/wiki/Roulette)
you can bet on several things including black or red. On this bet, if you win, you double your earnings. How does the casino make money on this then? If you look at the [possibilities](http://www.math.uah.edu/stat/games/Roulette.png)
you realize that the chance of red or black are both slightly less than 1/2. There are two green spots, so the of landing on black (or red) is actually 18/38, or 9/19.


## Problem 1A

Let's make a quick sampling model for this simple version of roulette. You are going to bet a dollar each time you play and always bet on black. Make a box model for this process using the `sample` function. Write a function `get_outcome` that takes as an argument the number of times you play $N$ and returns your earnings $S_N$.

```{r}

> get_outcome <- function(N) {
+          #N is number of spins of the wheel. Probabilities as stated. Outcomes are therefore
+                   results <- sample( c(1,-1), N, replace=TRUE, prob=c(9/19,10/19))
+                   # total earnings are
+                           S_N <- sum(results)
+                         avg <- S_N/N
+                          list(spins=N, results=results, tot_earnings=S_N, avg_earn_spin=avg)
+                           }
> get_outcome(5)
> # NOTE: please remove 'results' listing if there are a large number of input spins!
```

## Problem 1B

Use Monte Carlo simulation to study the distribution of total earnings $S_N$ for $N=10,25,100,1000$. That is, study the distribution of earnings for different number of plays. What are the distributions of these two random variables? How do the expected values and standard errors change with $N$? Then do the same thing for the average winnings $S_N/N$. What result that you learned in class predicts this?

```{r}
> # reviewing the distributions of both total winnings and average winnings given N spins
> # so will examine the mean and standard error of both variables for N spins
> # via Monte Carlo simulation (using 10,000 replications of N spins)

> # constant values
> mu <- -1/19
> sigma <- 2*sqrt(9/19*10/19)

> # Monte Carlo simulation for TOTAL earnings as a function of N spins
> 
> # code shown for 10 spins; repeated with values of N=25, 100, 1000 (only partial code shown for other values)
> spins <- 10
>  values_tot <- replicate (10^4, {
+          X <- sample( c(1,-1), spins, replace=TRUE, prob=(c(9/19,10/19))   )
+          sum(X)
+      })
> mean(values_tot)
[1] -0.455
> sample_se <- sqrt(spins)*sigma
> sample_se
[1] 3.157895

> spins <- 25
> mean(values_tot)
[1] -1.3044
> sample_se <- sqrt(spins)*sigma
> sample_se
[1] 4.99307

> spins <- 100
> values_tot <- replicate (10^4, {
+                       X <- sample( c(1,-1), spins, replace=TRUE, prob=(c(9/19,10/19))   )
+                      sum(X)
+                  })
> mean(values_tot)
[1] -5.1748
> sample_se <- sqrt(spins)*sigma
> sample_se
[1] 9.98614

> spins <- 1000
> mean(values_tot)
[1] -52.6252
> sample_se <- sqrt(spins)*sigma
> sample_se
[1] 31.57895

> # As this game is played more, the negative expected payoff per spin, -1/19 dollars, is magnified as you play. Expected losses grow until you stop playing. 
> # The standard error also grows as the range of possible results increases once you have had more time to play.


> # Monte Carlo simulation for AVERAGE earnings as a function of N spins

> # code shown for 10 spins; repeated with values of N=25, 100, 1000 (only partial code shown for other values)

> spins <- 10
> values_avg <- replicate (10^4, {
+                                                   X <- sample( c(1,-1), spins, replace=TRUE, prob=(c(9/19,10/19))   )
+                                                   mean(X) - mu
+                                            })
> mean(values_avg)
[1] -0.004928421
> sample_se <- sigma/sqrt(spins)
> sample_se
[1] 0.3157895

> spins <- 25
> mean(values_avg)
[1] -0.0001844211
> sample_se <- sigma/sqrt(spins)
> sample_se
[1] 0.1997228

> spins <- 100
> mean(values_avg)
[1] 0.0005415789
> sample_se <- sigma/sqrt(spins)
> sample_se
[1] 0.0998614

> spins <- 1000
> mean(values_avg)
[1] 0.0002347789
> sample_se <- sigma/sqrt(spins)
> sample_se
[1] 0.03157895

> # as the number of spins increases, the average payoff at the end end is closer to the "true" payoff of $ -1/19 given the odds of the game.
> # The standard error of the sample mean decreases as the sample mean gets closer to the true mean when you play more.
```

(note I also put some comments in the code)

The results we see here for the average payoff conform to the law of large numbers as there are more spins and average payoff after playing the game gets closer to the expected payoff under the (assumed) exact properties of the roulette wheel. 


## Problem 1C

What is the expected value of our sampling model? What is the standard deviation of our sampling model?

The expected value of payoff to the player is [9/19*(1) + 10/19*(-1)] or -1/19 dollars each time they play.
Because the outcomes are -1 or 1 dollar change in net position, the standard deviation for the binomial distribution 
here is 2*sqrt(spins)*sqrt[(10/19)*(9/19)] or about 0.998613998*sqrt(spins) for N number of spins.


## Problem 1D

Use CLT to approximate the probability that the casino loses money when you play 25 times. Then use a Monte Carlo simulation to confirm.

```{r}
> # The casino loses money when you win, and they have 9/19 chance of losing and 10/19 chance of winning in a spin.
> # Question is same as what are the odds you come out ahead. Same as what are chances you gross $13 or more/ win 13+ times.
> 
> # We know sigma is 2*sqrt(9/19*10/19). Standard error is this amount times sqrt(25), or 5, so 10*sqrt(9/19*10/19).

> # Using CLT, can develop a Z score. Want probability that you win at least 13 times. 
> # Z = [13 - (9/19)*25]  /   10*sqrt(9/19*10/19)   is the Z score distributed ~N(0,1)

> top <-  (13 - ((9*25)/19))
> top
[1] 1.157895
> bottom <- (10)*sqrt(9/19*10/19)
> bottom
[1] 0.1997228
> Z <- top / bottom
> Z
[1] 0.2319004
> pnorm(Z)
[1] 0.5916923

> # So there is about a 40% chance you can do as least as well as 13 games and cause a loss for the casino.
> # In other words, with 25 spins it is a 40% chance the casino would take a loss. 



> # generate a Monte Carlo sample and compare to the Z scores
error <- replicate(B,{
+          X <- sample( c(-1,1), N, replace=TRUE, prob=(c(10/19,9/19)) )
+          sum(X)
+      })

> # SUM above are your net gains, what is probability they are positive?
> mean (error > 0)
[1] 0.3971

> # This roughly matches the CLT computations above.


```



## Problem 1E

In general, what is the probability that the casino loses money as a function of $N$? Make a plot for values ranging from 25 to 1,000. Why does the casino give you free drinks if you keep playing?

```{r}
> # As noted, Pr(casino loses) = Pr(player makes money)

> DATA <- NULL

> for(spins in 25:1000) {
+                   i <- spins - 24 
+         winnings <- replicate(10^4,{
+             X <- sample( c(-1,1), spins, replace=TRUE, prob=(c(10/19,9/19)))
+                     sum(X)
+     })
+         prob <- mean(winnings > 0)  
+         DATA[i] <- prob 
+          }

> plot(DATA, xlab="spins",ylab="Pr(Casino loses)")
```

DATA contains the probability the casino will lose money given the number of spins of the wheel in a series of plays. It follows a curve decreasing at a decreasing rate. (note: am having trouble pasting in the graph, so will describe).

At low N, you can see that there is more chance for player to make money (and casino to lose it) when there is an odd number
of spins in the game. It is less noticeable for large N, where there is convergence to 0 (casino will almost surely win if the
player keeps on playing - hence the free drinks, comped room, etc.)
 

# Problem 2 

You run a bank that has a history of identifying potential homeowners that can be trusted to make payments. In fact, historically, in a given year, only 2% of your customers default. You want to use stochastic models to get an idea of what interest rates you should charge to guarantee a profit this upcoming year. 

## Problem 2A

Your bank gives out 1,000 loans this year. Create a sampling model and use the function `sample` to simulate the number of foreclosure in a year with the information that 2% of customers default. Also suppose your bank loses $120,000 on each foreclosure. Run the simulation for one year and report your loss.

```{r}
> # Doing a sample model for a single year of loans, modelling a loss of $120k with 1/50 probability, else no loss
 
> loans <- 1000

> outcome <- replicate (1, {
+      L <- sample (c(0,-120000), loans, replace=TRUE, prob=(c(49/50,1/50))  )
+       sum(L)
+      }  )
> outcome
[1] -2520000
> # The above single run of code showed the bank would lose $2.52 million from default

> # This result is fairly close to the expected. More replications should show an even closer average:
> outcome <- replicate (10^4, {
+ L <- sample (c(0,-120000), loans, replace=TRUE, prob=(c(49/50,1/50))  )
+ sum(L)
+ }   )
> mean(outcome)
[1] -2398992
```

## Problem 2B

Note that the loss you will incur is a random variable. Use Monte Carlo simulation to estimate the distribution of this random variable. Use summaries and visualization to describe your potential losses to your board of trustees.

```{r}
> # Looking at it as binomial distribution, you can draw 1 (a $120k loss) with Pr=.02 or else 0 (no loss) with probability .98
> # Expected # losses where n = 1000 is  np  =  1000 * .02  = 20 losses. Value of expected losses = $120k * 20 = $2.4 million
> # Standard error for the draw is sqrt(1000*.98*.02) = 4.427  Value would be 4.427 * $120K or ~$532k
 
> # Question 2a with 10k replications showed about a $2.4 million loss, close to expected value.

> #For standard error, will compute directly via MC method, approximating the value of square root of variance
> #Variance of # losses =  [E(L^2) - ([E(L)]^2)]

> # Compute E(L^2)
> L <- replicate (10^4, {
+                                                                 X <- sample( c(-120000,0), 1000, replace=TRUE, prob=(c(1/50,49/50))   )
+                                                                 sum(X)^2
+                                                           })
> x <- mean(L)
> x
[1] 6.089469e+12

> # Compute [E(loss)]^2
>  L2 <- replicate (10^4, {
+                                                                      X <- sample( c(-120000,0), 1000, replace=TRUE, prob=(c(1/50,49/50))   )
+                                                                      sum(X)
+                                                                })
> y <- mean(L2)^2
> y
[1] 5.728823e+12

> sample_var <- x-y
> sample_var
[1] 3.60646e+11
> sample_se <- sqrt(sample_var)
> sample_se
[1] 600538.1
> # Approximation is decent... ~$600k standard error.

> # keep simple labels for the executives
> # make a continuous plot

> values <- replicate (10^4, {
+                                                                                           X <- sample( c(-120000,0), 1000, replace=TRUE, prob=(c(1/50,49/50))   )
+                                                                                            sum(X)
+                                                                                   })
> values_mil <- values / 1000000

> options (scipen=5)
> d <- density(values_mil)
> medx <- median(values_mil)

> # THIS MAKES THE GRAPH
> plot(d, xlab="expected profit in millions of $", ylab="likelihood", main="Potential Negative Profits from Loan Default: red line splits values 50/50")
> abline(v = medx, col = "red", lwd = 2)


> # Graph is a quick idea of the potential loss to the bank. Should explain that although the average historical loss is $2.4
> # million, it can be 0 at best(which is rare) and can also be very steep (also rare,  $120,000 * 1000  is the worst)

```

## Problem 2C

The 1,000 loans you gave out were for $180,000. The way your bank can give out loans and not lose money is by charging an interest rate. If you charge an interest rate of, say, 2% you would earn $3,600 for each loan that doesn't foreclose. At what percentage should you set the interest rate so that your expected profit totals $100,000. Hint: Create a sampling model with expected value 100 so that when multiplied by the 1,000 loans you get an expectation of $100,000. Corroborate your answer with a Monte Carlo simulation.

From before, we know that the expected loss is $2.4 million. To expect to make $100k, the bank needs to make $2.5 million off the
980 loans not defaulting. $2.5MM/980 = $2551 profit per good loan. Since the loans are for $180k, this implies an interest rate of
$2551/$180k, about 1.4%.


```{r}
> # For this part I write a function that uses the given default probabilities, and computes the expected return for an input interest rate.
> # Method is to start from 2% and grid search in direction until inputting an interest rate gives an average profit per outstanding loan of ~$100.
> # (I admit the answer I just wrote is telling me what direction to go in!)
> # Note that although there are 1000 loans, since it is average I am looking at, I used 10^5 as series argument to get more precision for the average.

> outcome <- function(int) {
+                                              profit <- sample( c(int*180000,-120000), 10^5, replace=TRUE, prob=c(49/50,1/50))
+                                                             
+                                                                                                  sum <- sum(profit)
+                                                                                                        avg <- sum/10^5
+                                                                                                                               list(avg_profit=avg)
+                                                                                                                               }
> outcome(.02)
$avg_profit
[1] 1167.552

> outcome(.01)
$avg_profit
[1] -553.176

> outcome(.015)
$avg_profit
[1] 217.779

> outcome(.014)
$avg_profit
[1] 68.3748

> outcome(.0145)
$avg_profit
[1] 182.322

> outcome(.0142)
$avg_profit
[1] 179.6392

> outcome(.0141)
$avg_profit
[1] 7.5903

> #With a rate of 1.42% (rough estimate), the expected return on a good loan is .0142*180k or $2556
> # Confirm with MC method by using this return and looking at the expected profit
> profit <- replicate(10^4,{
+                         P <- sample( c(-120000,2556), 1000, replace=TRUE, prob=(c(1/50,49/50)) )
+                         mean(P)
+                     })
> mean(profit)
[1] 107.5517
> # This is appropriately close to $100 per loan for the 1,000 loans.
```


## Problem 2D

In problem 2C, you were able to set a very low interest rate. Your customers will be very happy and you are expected to earn $100,000 in profits. However, that is just an expectation. Our profit is a random variable. If instead of a profit your bank loses money, your bank defaults. Under the conditions of Problem 2C, what is the probability that your profit is less than 0?

```{r}
> # Using the model we just made, see what is the probability that the average (or total, same idea) is less than zero?

>  profit <- replicate(10^4,{
+                              P <- sample( c(-120000,2556), 1000, replace=TRUE, prob=(c(1/50,49/50)) )
+                              mean(P)
+                          })
> 
> mean(profit < 0)
[1] 0.4357

> # This is quite a high chance, 44%, the bank may not be comfortable with that.
```

## Problem 2E

Note that the probability of losing money is quite high. To what value would you have to raise interest rates in order to make the probability of losing money, and your bank and your job, as low as 0.001? What is the expected profit with this interest rate? Corroborate your answer with a Monte Carlo simulation.

Hint: Use the following short cut. If $p$ fraction of a box are $a$s and $(1-p)$ are $b$s, then the SD of the list is $\mid a-b \mid \sqrt{p(1-p)}$ 

(I admit I was slowly getting here ... thanks for the boost in the GitHub discussion!)

As noted in the discussion, the interest rate "int" is not known and needs to be set. Expected profits over the loans are:
EP = 1000  * (  [180k*int]*49/50  +  [-120k]*1/50  )
SE = SD*sqrt(1000)   =   sqrt(1000)*abs(180k*int + 120k)*sqrt(1/50*49/50)

Using CLT, we can get a Z score for profit level where .001 of the area is to the left of the Z score corresponding to zero
profits. That Z score is -3.09. What interest rate goes with that? Profit at Z=-3.09 is 0, so Z score = 0-EP/SE as above.
Solving for -EP/SE = -3.09 gives int = 2.32%. Plugging into expected profit equation yields EP just under $1.7 million.



```{r}
> # With interest rate = 2.32% the bank makes about $4176 (.0232*$180k) off the loans that dont go bad. 
> # Monte Carlo profit simulation:

> profit <- replicate(10^4,{
+                                       P <- sample( c(-120000,4176), 1000, replace=TRUE, prob=(c(1/50,49/50)) )
+                                        sum(P)
+                                    })
> mean(profit)
[1] 1698267

> mean(profit < 0)
[1] 0.0023

> # mean is close to predicted but the probability of loss is a bit higher
```

## Problem 2F

Note that the Monte Carlo simulation gave a slightly higher probability than 0.001. What is a possible reason for this? 
Hint: See if the disparity is smaller for larger values of $p$. Also check for probabilities larger than 0.001. Recall we made an assumption when we calculated the interest rate.


```{r}
##your code here
```


Your answer here.

## Problem 3

We were able to set an interest rate of about 2% that guaranteed a very low probability of having a loss. Furthermore, the expected average was over $1 million. Now other financial companies noticed the success of our business. They also noted that if we increase the number of loans we give, our profits increase. However, the pool of reliable borrowers was limited. So these other companies decided to give loans to less reliable borrowers but at a higher rate.

## Problem 3A

The pool of borrowers they found had a much higher default rate, estimated to be $p=0.05$. What interest rate would give these companies the same expected profit as your bank (Answer to 2E)? 

```{r}
##your code here
```

## Problem 3B 

At the interest rate calculated in 3A what is the probability of negative profits? Use both the normal approximation and then confirm with a Monte Carlo simulation.

```{r}
##your code here
```

## Problem 3C 

Note that the probability is much higher now. This is because the standard deviation grew. The companies giving out the loans did not want to raise interest rates much more since it would drive away clients. Instead they used a statistical approach. They increased $N$. How large does $N$ need to be for this probability to be 0.001? Use the central limit approximation and then confirm with a Monte Carlo simulation.

Your answer here.
```{r}
###your code here
```

So by doubling the number of loans we were able to reduce our risk! Now, for this to work, all the assumptions in our model need to be approximately correct, including the assumption that the probability of default was **independent**. This turned out to be false and the main reason for the under estimation of risk.


## Problem 3D

Define the following matrix of outcomes for two borrowers using our previous box model:

```{r}
loan <- 180000
loss_per_foreclosure <- 120000
p2 <- 0.05
interest_rate2 <- 0.05
B <- 10^5
outcomes1 <- replicate(B,{
  sample( c(-loss_per_foreclosure, interest_rate2*loan ), 2, replace=TRUE, prob=c(p2, 1-p2))
})
```
We can confirm independence by computing the probability of default for the second conditioned on the first defaulting: 

```{r}
sum( outcomes1[1,] < 0 & outcomes1[2,]<0)/sum(outcomes1[1,]<0)
```

This quantity is about the same as the probability of default $0.05$.

Now we create a new model. Before generating each set of defaults, we assume that a random event occurred that makes all default probabilities go up or go down by 4 points. We could see how this would happen if, for example, demand for houses decreases and all house prices drop. 

```{r}
B <- 10^5
outcomes2 <- replicate(B,{
  add <- sample( c(-0.04,0.04), 1)
  sample( c(-loss_per_foreclosure, interest_rate2*loan ), 2, replace=TRUE, prob=c(p2+add, 1-(p2+add)))
})
```

Note that the outcomes are no longer independent as demonstrated by this result not being equal to 0.05

```{r}
sum( outcomes2[1,] < 0 & outcomes2[2,]<0)/sum(outcomes2[1,]<0)
```


Generate a simulation with correlated outcomes such as those above. This time use the interest rate calculated in 3A. What is the expected earnings under this model compared to the previous? What is the probability of losing $1 million compared to the previous? What is the probability of losing $10 million compared to the previous?



```{r}
###your code here
```


## Problem 4

Read [this wikipedia page](https://en.wikipedia.org/wiki/Financial_crisis_of_2007%E2%80%9308) about the financial crisis. Write a paragraph describing how what you learned in this homework can help explain the conditions that led to the crisis.
